Boosting Higgs Pair Production in the $bb\bar{b}\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-06 paper/sec-mva.tex

After the traditional cut-based analysis presented in the previous section,
and combining the significance of the different event categories,
we end up with a $S/\sqrt{B}$ value of AAA.
%
Taking into account our more realistic simulation of the QCD background as
compared to previous studies, this is not too bad.
%
However, the attentive reader might have noticed that we have used
relatively 
loose kinematics cuts, for example in the Higgs mass window.
%
The motivation for this is that the $n$-tuples of signal and background
events, after our basic selection, are processed though a multivariate
analysis with the motivation to increase the signal over background
significance.
%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis.

@@

The classification of events as arising from either signal or
background processes via the use of a multivariate analysis is a common
technique in high energy physics.

In order to investigate the potential efficacy of an MVA in an $HH\to 4b$
measurement, we process the events remaining after the cut-based analysis 
through a standard feed-forward artificial neural network with architecture

\be
\label{eq:nn1}
N_{\mathrm{var}}\times5\times3\times1 \, ,
\ee
where $N_{\mathrm{par}}$ represents the number of input variables for the MVA.
All neural-network layers use a sigmoid activation function, allowing for a probabilistic
interpretation of the neural-network output.

The input variables for the MVA differ between our three categories, based upon
the availability in each analysis of substructure information on fat jets. For all
three analyses, the following common variables are used:

\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs candidates, $p_{T,H_0}$, $p_{T,H_1}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,HH}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{H,1}$, $m_{H,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{HH}$.
\item The separation in $R$ of the two Higgs candidates, $\Delta R_{HH}$.
\item The separation in $\phi$ of the two Higgs candidates, $\Delta \phi_{HH}$.
\item The separation in $\eta$ of the two Higgs candidates, $\Delta \eta_{HH}$.
\item The transverse momenta of the leading Higgs candidate subjets, $p_{T,H_{0,0}}$, $p_{T,H_{0,1}}$
\item The transverse momenta of the sub-leading Higgs candidate subjets, $p_{T,H_{1,0}}$, $p_{T,H_{1,1}}$
\end{itemize}

For analyses including the reconstruction of fat jets, the following variables are included for
every fat jet Higgs candidate present in the analysis.

\begin{itemize}
\item The $k_T$ splitting scale of the fat-jet (Reference).
\item The ratio of 2-to-1-Subjetiness $\tau_{2}/\tau_{1}$ of the fat-jet (Reference).
\item The fat-jet energy correlation function double-ratio $C_2$ (Reference).
\item The fat-jet energy correlation function double-ratio $D_2$ (Reference).
\end{itemize}

Therefore there are in total $N_{\mathrm{var}}=13$ variables for the resolved analysis, 17 for the semi-resolved, and 21 for the boosted category.

Given the above set of kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network parameters $\{\omega\}$, we interpret the neural network output $y_i$ as the probability that the event $i$ originates from the signal process,
\be y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} ), \ee
where $y_i^\prime$ represents the true classification of the event $i$, i.e $y^\prime_{\text{signal}} = 1$ and $y^\prime_{\text{background}} = 0$. With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i}.
\ee
 Correspondingly the negative log-likelihood cost function to be used in the neural network training is the cross-entropy error
\be E(\{\omega\}) = -\log\left(\prod_i^{N_{\text{evt}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right) = \sum_i^{N_{\text{evt}}} y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)}. \ee

Using the cross-entropy cost function, the neural networks are trained with a basic genetic algorithm (GA) over 50,000 generations.
(Do we need to specify GA params etc?).
\subsection{Enhancing signal discrimination with the MVA}

@@

The corresponding analysis for the boosted category is also
shown in Fig.~\ref{fig:nnweights}.
%
Here we see that there is a more even distributions of weights, in that many
variables help in a similar amount in the discrimination.
%
The invariant mass of the Higgs candidate of course still carries a lot of weight,
but now we see that the substructure variables really carry  a lot of information,
in particular the splitting scales and the subjetiness $\tau_{12}$.
%
We have verified that signal significance in the boosted category is substantially
degraded if substructure variables are not used.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/tau21_H0_res_C1_boost.pdf}
  \includegraphics[width=0.48\textwidth]{plots/EEC_C2_H0_res_C1_boost.pdf}
  \caption{\small Left plot: subjetiness.
    Right plot: energy correlations.
}
\label{fig:mva_substructure_1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate this, we now consider a variant of the analysis used in
Table  where the $m_H$ window criterion has been relaxed to $80 < m_H < 170$.
%
Firstly the same analysis is repeated with the larger $m_H$ window, with no smearing at all.
%
This is then compared to the same analysis but with a Gaussian smearing with $\sigma=10$ GeV applied to the dijet invariant mass used in the MVA.
%
In the left panel of Fig.~\ref{fig:invMsmear} the smearing is demonstrated.
%
Both with and without smearing, the  invariant mass distribution of the
Higgs candidates is still the most discriminative variable.
%
This can be seen by comparing with   with the corresponding distribution with smearing in
  Fig.~\ref{fig:nnweights}.

>>>>

After the traditional cut-based analysis presented in the previous section,
and combining the significance of the different event categories,
we end up with a $S/\sqrt{B}$ value of AAA.
%
Taking into account our more realistic simulation of the QCD background as
compared to previous studies, this is not too bad.
%
however, the attentive reader might have noticed that we have used
relatively 
loose kinematics cuts, for example in the Higgs mass window.
%
The motivation for this is that the $n$-tuples of signal and background
events, after our basic selection, are processed though a multivariate
analysis with the motivation to increase the signal over background
significance.
%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis.

@@

The classification of events as arising from either signal or
background processes via the use of a multivariate analysis is a common
technique in high energy physics.
%
In order to investigate the potential efficacy of an MVA in an $hh\to 4b$
measurement, we process the events remaining after the cut-based analysis 
through a feed-forward artificial neural network with architecture
\be
\label{eq:nn1}
N_{\mathrm{var}}\times5\times3\times1 \, ,
\ee
where $N_{\mathrm{var}}$ represents the number of input variables for the MVA.
All neural-network layers use a sigmoid activation function, allowing
for a probabilistic
interpretation of the neural-network output.

Given a set of $N_{\mathrm{var}}$  kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network parameters $\{\omega\}$, we interpret the neural network output $y_i$ as the probability that the event $i$ originates from the signal process,
\be y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} ), \ee
where $y_i^\prime$ represents the true classification of the event $i$, i.e $y^\prime_{\text{signal}} = 1$ and $y^\prime_{\text{background}} = 0$. With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i}.
\ee
 Correspondingly the negative log-likelihood cost function to be used in the neural network training is the cross-entropy error, defined as
 \bea
 E(\{\omega\}) &=& -\log\left(\prod_i^{N_{\text{evt}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right)\nonumber\\
 &=&
 \sum_i^{N_{\text{evt}}} y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)} \, ,
 \label{cross-entropy}
 \eea
 where $N_{\text{evt}}$ is the number of
 Monte Carlo events that we have generated.

 The training of the neural networks consist on the
 minimization of the cross-entropy error,
 Eq.~\ref{cross-entropy}, which is achieved using a
 Genetic Algorithm.
 %
 Since given the large number of events the probability of
 over-fitting is tiny, we just train the MVA for a very long
 number of generations, in this case $N_{\rm gen}=50k$.
 %
 We have verified that if a larger number of generations
 are used the results are unchanged.

 \subsection{MVA input}

 To identify the most discriminatory variables, we have to train
 the MVA using a large number of kinematic distributions for
 signal and background events.
%
In our $hh\to 4b$ case,
the input variables for the MVA differ between our three categories,
since in the case of large-$R$ jet reconstruction we want to exploit
the richness of substructure information.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs candidates, $p_{T,h_1}$, $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$, $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in $R$ of the two Higgs candidates, $\Delta R_{hh}$.
\item The separation in $\phi$ of the two Higgs candidates, $\Delta \phi_{hh}$.
\item The separation in $\eta$ of the two Higgs candidates, $\Delta \eta_{hh}$.
\item The transverse momenta of the leading Higgs candidate subjets, $p_{T,h_{1,1}}$, $p_{T,h_{1,2}}$
\item The transverse momenta of the sub-leading Higgs candidate subjets, $p_{T,h_{2,1}}$, $p_{T,h_{2,1}}$
\end{itemize}
Note that in the case of a large-$R$ jet, the subjets are defined
with the anti-$k_t$ algorithm with $R=0.3$.
%
In the resolved case, the subjets are instead replaced by the two
small-$R$ jets that constitute each Higgs candidate.

For analyses involving the use of large-$R$ jets,
namely for the boosted and intermediate categories,
the following variables are included for
every large-$R$ jet,
we use in addition a number of substructure variables~\cite{Aad:2013gja},
in particular:
\begin{itemize}
\item The $k_T$ splitting scale~\cite{Butterworth:2008iy}
\item The ratio of 2-to-1-Subjetiness $\tau_{2}/\tau_{1}$~\cite{Thaler:2010tr}
\item The energy correlation function double-ratio $C_2$~\cite{Larkoski:2013eya}
\item The energy correlation function double-ratio $D_2$~\cite{Larkoski:2013eya}
\end{itemize}
The implementation of all these substructure variables follows
closely the original ATLAS implementation summarized in Ref.~\cite{Aad:2013gja}.

Therefore, there are in total $N_{\mathrm{var}}=13$ variables for the resolved analysis, 17 for the semi-resolved, and 21 for the boosted category.
%
Given that the MVA is able to identify the most discriminatory variables
and to ignore those that have smaller effect, it is advantageous to
include as many substructure variables as possible.

\subsection{Enhancing signal discrimination with the MVA}

@@

The corresponding analysis for the boosted category is also
shown in Fig.~\ref{fig:nnweights}.
%
here we see that there is a more even distributions of weights, in that many
variables help in a similar amount in the discrimination.
%
The invariant mass of the Higgs candidate of course still carries a lot of weight,
but now we see that the substructure variables really carry  a lot of information,
in particular the splitting scales and the subjetiness $\tau_{12}$.
%
We have verified that signal significance in the boosted category is substantially
degraded if substructure variables are not used.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/tau21_h0_res_C1_boost.pdf}
  \includegraphics[width=0.48\textwidth]{plots/EEC_C2_h0_res_C1_boost.pdf}
  \caption{\small Left plot: subjetiness.
    Right plot: energy correlations.
}
\label{fig:mva_substructure_1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate this, we now consider a variant of the analysis used in
Table  where the $m_h$ window criterion has been relaxed to $80 < m_h < 170$.
%
Firstly the same analysis is repeated with the larger $m_h$ window, with no smearing at all.
%
This is then compared to the same analysis but with a Gaussian smearing with $\sigma=10$ GeV applied to the dijet invariant mass used in the MVA.
%
In the left panel of Fig.~\ref{fig:invMsmear} the smearing is demonstrated.
%
Both with and without smearing, the  invariant mass distribution of the
Higgs candidates is still the most discriminative variable.
%
This can be seen by comparing with   with the corresponding distribution with smearing in
  Fig.~\ref{fig:nnweights}.
